---
title: "Project_part1_atreml"
author: "Alexandra Treml"
date: "2024-07-20"
output: html_document
---
Part 1 includes: 
1. set up git
2. read in the data from genes and series matrix
3. id 1 gene, 1 continuous variable, and 2 categorical covariates in the provided datasets (must link gene expression data and metadata)
4. Generate 3 plots using ggplot2 for your covariates of choice
  a. hist for gene expression
  b. scatterplot for gene expression and continuous covariate
  c. boxplot of gene expression separated by both categorical variables
  
  #should filter out ICU from Non-ICU


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## R Markdown

```{r}
# Set the working directory and read in the files
setwd("C:/Users/AlexandraTreml/Desktop/MS/QBS103/Project/QBS103_proj")
genes <- read.csv("genes_GSE157103.csv")
participant <- read.csv("series_matrix_GSE157103.csv")

head(genes)
head(participant)

# Rename the first column in genes to 'gene'
genes <- genes %>%
  rename(gene = X)

# Filter for 'A2M' gene
genes <- genes %>%
  filter(gene == "A2M")

# Pivot the genes dataframe
genes_long <- genes %>%
  tidyr::pivot_longer(cols = -gene, names_to = 'id', values_to = 'expression')
#head(genes_long)

# rename id column in participant, select my categorical covariates and continuous variable (+age for eda)
participant <- participant %>%
  filter(icu_status == ' yes') %>%
  rename(id = participant_id) %>%
  select(id, age, sex, mechanical_ventilation, lactate.mmol.l.)
#head(participant)

#join 2 dataframes
df <- left_join(participant, genes_long, by = "id")
# head(df)

#clean up new dataframe

#fill lactate unknowns with NA 
df$lactate.mmol.l.[df$lactate.mmol.l. == " unknown"] <- NA
#fill sex unknowns with NA 
df$sex[df$sex == " unknown"] <- NA
#remove : from age, and fill with NA
df$age[df$age == " :"] <- NA

head(df)
```
```{r}
#create a histogram using sex and gene expression

# Load the necessary library
library(ggplot2)
df <- na.omit(df)
df <- df %>% 
  select(id, gene, sex, age, mechanical_ventilation, expression, lactate.mmol.l.)
head(df)

df$sex <- as.factor(df$sex)
levels(df$sex)


ggplot(data = df, aes(x = expression, fill = sex)) + 
  geom_histogram(binwidth = 0.05, position = 'dodge') + 
  labs(x = 'Gene Expression', y = 'Frequency', title = 'A2M Gene Expression Among COVID ICU Patients') +  
  scale_fill_manual(values = c("violetred", "royalblue2"))

```
```{r}
#create a scatterplot using gene expression and lactate 

# Load necessary libraries
library(dplyr)
library(ggplot2)

# Convert lactate.mmol.l. to numeric
df$lactate.mmol.l. <- as.numeric(df$lactate.mmol.l.)

# Remove leading and trailing whitespaces from sex column
df$sex <- trimws(df$sex)

# Check unique levels
unique(df$sex)

# Set factor levels
df$sex <- factor(df$sex, levels = c("male", "female"))

# Filter out values greater than 6.5 for the y-axis
df <- df %>% filter(lactate.mmol.l. <= 6.5)

# Create the plot with the filtered data
ggplot(df, aes(x = expression, y = lactate.mmol.l., color = sex, shape = sex)) + 
  geom_point(size = 3) + 
  scale_y_continuous(breaks = seq(0, ceiling(max(df$lactate.mmol.l., na.rm = TRUE)), by = 0.5)) +
  geom_smooth(method = "lm", se = FALSE) + 
  labs(x = 'Gene Expression', y = 'Lactate (mmol/l)', title = 'Gene Expression vs Lactate for COVID ICU Patients') + 
  theme_classic(base_size = 5) +
  scale_color_manual(values = c("male" = "royalblue2", "female" = "violetred3")) + 
  scale_shape_manual(values = c("male" = 16, "female" = 17)) +  
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"), # Center and bold title
    axis.title = element_text(size = 14, face = "bold"), # Bold axis titles
    axis.text = element_text(size = 12), # Increase axis text size
    legend.title = element_text(size = 12), 
    legend.text = element_text(size = 12) 
  )

```

```{r}
#boxplot of gene expression separated by both ventilator, sex, and age group

head(df)

df$age <- as.numeric(df$age)

df$age_group <- cut(df$age,
                    breaks = c(21, 30, 50, 70, 88),
                    labels = c("21-30", "31-50", "51-70", "71-88"),
                    right = TRUE)
head(df)


ggplot(df, aes(x = mechanical_ventilation, y = expression, color = sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("male" = "royalblue1", "female" = "violetred1")) +
  labs(x = 'Mechanical Ventilation', y = 'Gene Expression', title = 'A2M Gene Expression Among Ventilated vs Non-Ventilated COVID ICU Patients') +
  theme_classic()

ggplot(df, aes(x = age_group, y = expression, color = sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("male" = "royalblue1", "female" = "violetred1")) +
  labs(x = 'Age Group', y = 'Gene Expression', title = 'A2M Gene Expression Among COVID ICU Patient Age Groups') +
  theme_classic()
```




