---
title: "Project_part2_atreml"
author: "Alexandra Treml"
date: "2024-08-06"
output:
  pdf_document: default
---
Part 2 includes: 
1. Build a function to create the plots from part 1.
Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates

2. Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created

3. Present one of your boxplots in class.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## R Markdown

```{r}
#DATA CLEANING AND PREP

# Set the working directory and read in the files
setwd("C:/Users/AlexandraTreml/Desktop/MS/QBS103/Project/QBS103_proj")
genes <- read.csv("genes_GSE157103.csv")
participant <- read.csv("series_matrix_GSE157103.csv")

# Rename the first column in genes to 'gene'
genes <- genes %>%
  rename(gene = X)

# Pivot the genes dataframe
genes_long <- genes %>%
  tidyr::pivot_longer(cols = -gene, names_to = 'id', values_to = 'expression')
#head(genes_long)

# rename id column in participant, select my categorical covariates and continuous variable (+age for eda)
participant <- participant %>%
  filter(icu_status == ' yes') %>%
  rename(id = participant_id) %>%
  select(id, age, sex, mechanical_ventilation, lactate.mmol.l.)
#head(participant)

#join 2 dataframes
df <- left_join(participant, genes_long, by = "id")

#clean up new dataframe

#fill lactate unknowns with NA 
df$lactate.mmol.l.[df$lactate.mmol.l. == " unknown"] <- NA
#fill sex unknowns with NA 
df$sex[df$sex == " unknown"] <- NA
#remove : from age, and fill with NA
df$age[df$age == " :"] <- NA

head(df)
```
```{r}
library(ggplot2)
library(dplyr)

hist_func <- function(df, gene_name, cont_covar, cat_covar1, cat_covar2) {
  df <- na.omit(df)
  
  #convert to symbols to use in plotting - unsure how to correct this
  cont_covar <- sym(cont_covar)
  cat_covar1 <- sym(cat_covar1)
  cat_covar2 <- sym(cat_covar2)
  
  df <- df %>% 
    filter(gene == gene_name) %>%
    select(id, gene, !!cont_covar, !!cat_covar1, !!cat_covar2)
  
  df[[as.character(cat_covar1)]] <- as.factor(df[[as.character(cat_covar1)]])
  
  ggplot(data = df, aes(x = !!cont_covar, fill = !!cat_covar2)) + 
    geom_histogram(binwidth = 0.05, position = 'dodge') + 
    labs(x = 'Gene Expression', y = 'Frequency', 
         title = paste(gene_name, 'Gene Expression Among COVID ICU Patients')) +  
    scale_fill_manual(values = c("violetred", "royalblue2"))
}

# Call the function with column names as strings
hist_func(df, 'A1CF', 'expression', 'age', 'sex')
hist_func(df, 'A2M', 'expression', 'age', 'sex')
hist_func(df, 'A4GALT', 'expression', 'age', 'sex')

```


```{r}
#create a scatterplot using gene expression and lactate 

# Load necessary libraries
# Load necessary libraries
library(dplyr)
library(ggplot2)

scatterplot_func <- function(df, gene_name, expression, cont_covar, cat_covar1, cat_covar2) {
  
  # Filter the dataframe based on the gene_name argument
  df <- df %>% filter(gene == gene_name)
  
  # Convert columns to appropriate types
  df[[cont_covar]] <- as.numeric(as.character(df[[cont_covar]]))
  df[[cat_covar1]] <- trimws(as.character(df[[cat_covar1]]))
  df[[cat_covar2]] <- trimws(as.character(df[[cat_covar2]]))
  
  # Convert categorical columns to factors
  df[[cat_covar2]] <- factor(df[[cat_covar2]], levels = c("male", "female"))
  df[[cat_covar1]] <- factor(df[[cat_covar1]], levels = c("yes", "no"))
  
  # Filter based on conditions
  #df <- df %>% filter(df[[cont_covar]] <= 6.5, df[[expression]] <= 0.5)
  
  # Scatterplot
  ggplot(df, aes_string(x = expression, y = cont_covar, color = cat_covar1, shape = cat_covar1)) + 
    geom_point(size = 3) + 
    scale_y_continuous(breaks = seq(0, ceiling(max(df[[cont_covar]], na.rm = TRUE)), by = 0.5)) +
    geom_smooth(method = "lm", se = TRUE) + 
    labs(x = 'Gene Expression', y = 'Lactate (mmol/l)', title = paste(gene_name, 'Gene Expression vs Lactate for COVID ICU Patients')) + 
    theme_classic(base_size = 5) +
    scale_color_brewer(palette = 'Dark2') +
    scale_shape_manual(values = c(16, 17)) +  
    theme(
      plot.title = element_text(size = 16, face = "bold"), # Center and bold title
      axis.title = element_text(size = 12, face = "bold"), # Bold axis titles
      axis.text = element_text(size = 12), # Increase axis text size
      legend.title = element_text(size = 12), 
      legend.text = element_text(size = 12) 
    )
}

# Example calls to the function
scatterplot_func(df, 'A4GALT', 'expression', 'lactate.mmol.l.', 'mechanical_ventilation', 'sex')
scatterplot_func(df, 'A2M', 'expression', 'lactate.mmol.l.', 'mechanical_ventilation', 'sex')
scatterplot_func(df, 'A1CF', 'expression', 'lactate.mmol.l.', 'mechanical_ventilation', 'sex')

```

```{r}
#boxplot of gene expression separated by both ventilator, sex, and age group

scatterplot_func <- function(df, gene_name, expression, cont_covar, cat_covar1, cat_covar2) {
  
    df <- df %>% 
    filter(gene == gene_name)


    df$age <- as.numeric(df$age)

    df$age_group <- cut(df$age,
                     breaks = c(21, 30, 50, 70, 88),
                    labels = c("21-30", "31-50", "51-70", "71-88"),
                    right = TRUE)


ggplot(df, aes(x = mechanical_ventilation, y = expression, color = sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("male" = "royalblue1", "female" = "violetred1")) +
  labs(x = 'Mechanical Ventilation', y = 'Gene Expression', title = 'A2M Gene Expression Among Ventilated vs Non-Ventilated COVID ICU Patients') +
  theme_classic()

ggplot(df, aes(x = age_group, y = expression, color = sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("male" = "royalblue1", "female" = "violetred1")) +
  labs(x = 'Age Group', y = 'Gene Expression', title = 'A2M Gene Expression Among COVID ICU Patient Age Groups') +
  theme_classic()
} 
```




