---
title: "Project_part2_atreml"
author: "Alexandra Treml"
date: "2024-08-06"
output:
  pdf_document: default
---
Part 2 includes: 
1. Build a function to create the plots from part 1.
Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates

2. Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created

3. Present one of your boxplots in class.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

## R Markdown

```{r}
#DATA CLEANING AND PREP

# Set the working directory and read in the files
setwd("C:/Users/AlexandraTreml/Desktop/MS/QBS103/Project/QBS103_proj")
genes <- read.csv("genes_GSE157103.csv")
participant <- read.csv("series_matrix_GSE157103.csv")

# Rename the first column in genes to 'gene'
genes <- genes %>%
  rename(gene = X)

# Pivot the genes dataframe
genes_long <- genes %>%
  tidyr::pivot_longer(cols = -gene, names_to = 'id', values_to = 'expression')

#rename id column in participant, select my categorical covariates and continuous variable (+age for eda)
participant <- participant %>%
  filter(icu_status == ' yes') %>%
  rename(id = participant_id) %>%
  select(id, age, sex, mechanical_ventilation, lactate.mmol.l.)

#join 2 dataframes
df <- left_join(participant, genes_long, by = "id")

#fill lactate unknowns with NA 
df$lactate.mmol.l.[df$lactate.mmol.l. == " unknown"] <- NA
#fill sex unknowns with NA 
df$sex[df$sex == " unknown"] <- NA
#remove : from age, and fill with NA
df$age[df$age == " :"] <- NA

head(df)
```

```{r}
hist_func <- function(df, gene_name, cont_covar, cat_covar1, cat_covar2) {
  df <- na.omit(df) #get rid of na
  
  df <- df %>% 
    filter(gene == gene_name) %>% #filter for my gene in the for loop 
    select(id, gene, cont_covar, cat_covar1, cat_covar2) #select just the cols I want from df
  
  df[[cat_covar1]] <- as.factor(df[[cat_covar1]]) #create as factor for my plot
  
  p <- ggplot(data = df, aes_string(x = cont_covar, fill = cat_covar2)) + 
    geom_histogram(binwidth = 0.05, position = 'dodge') + 
    labs(x = 'Gene Expression', y = 'Frequency', 
         title = paste(gene_name, 'Gene Expression Among COVID ICU Patients')) +  
    scale_fill_manual(values = c("violetred", "royalblue2"))
  
  print(p)
}

genes <- c('A4GALT', 'A2M', 'A1CF') #make a list of my 3 genes

# Loop through gene names to create 3 plots
for (gene in genes) {
  hist_func(df, gene, 'expression', 'age', 'sex') #loop through each gene and call func
}

```



```{r}
#create a scatterplot using gene expression and lactate 

# Load necessary libraries
# Load necessary libraries
library(dplyr)
library(ggplot2)

scatterplot_func <- function(df, gene_name, expression, cont_covar, cat_covar1, cat_covar2) {
  
  #filter for my selected gene in the for loop 
  df <- df %>% filter(gene == gene_name)
  
  ##onvert to numeric and clean up spaces
  df[[cont_covar]] <- as.numeric(as.character(df[[cont_covar]])) 
  df[[cat_covar1]] <- trimws(as.character(df[[cat_covar1]]))
  df[[cat_covar2]] <- trimws(as.character(df[[cat_covar2]]))
  
  # Convert categorical columns to factors
  df[[cat_covar2]] <- factor(df[[cat_covar2]], levels = c("male", "female"))
  df[[cat_covar1]] <- factor(df[[cat_covar1]], levels = c("yes", "no"))
  
  # Scatterplot
  p <- ggplot(df, aes_string(x = expression, y = cont_covar, color = cat_covar1, shape = cat_covar1)) + 
    geom_point(size = 3) + 
    scale_y_continuous(breaks = seq(0, ceiling(max(df[[cont_covar]], na.rm = TRUE)), by = 0.5)) +
    geom_smooth(method = "lm", se = TRUE) + 
    labs(x = 'Gene Expression', y = 'Lactate (mmol/l)', title = paste(gene_name, 'Gene Expression vs Lactate for COVID ICU Patients')) + 
    theme_classic(base_size = 5) +
    scale_color_brewer(palette = 'Dark2') +
    scale_shape_manual(values = c(16, 17)) +  
    theme(
      plot.title = element_text(size = 16, face = "bold"), # Center and bold title
      axis.title = element_text(size = 12, face = "bold"), # Bold axis titles
      axis.text = element_text(size = 12), # Increase axis text size
      legend.title = element_text(size = 12), 
      legend.text = element_text(size = 12) 
    )
  print(p) #print each plot now that it's in a function 
}

genes <- c('A4GALT', 'A2M', 'A1CF')

#loop through my gene names to create 3 plots
for (gene in genes) {
  scatterplot_func(df, gene, 'expression', 'lactate.mmol.l.', 'mechanical_ventilation', 'sex')
}

```

```{r}
#boxplot of gene expression separated by both ventilator, sex
boxplot_func <- function(df, gene_name, expression, cat_covar1, cat_covar2) {
    df <- na.omit(df)
    df <- df %>% 
    filter(gene == gene_name) #filter for just the gene in the loop 

    p <- ggplot(df, aes_string(x = cat_covar1, y = expression, color = cat_covar2)) +
      geom_boxplot() +
      scale_fill_manual(values = c("male" = "royalblue1", "female" = "violetred1")) +
      labs(x = cat_covar1, y = 'Gene Expression', title = paste(gene_name, 'Gene Expression Among Ventilated vs Non-Ventilated COVID ICU Patients')) +
      theme_classic()

    print(p) #print each plot now that it's a function
} 

genes <- c('A4GALT', 'A2M', 'A1CF') #set my 3 genes

# Loop through gene names to create 3 plots
for (gene in genes) {
  boxplot_func(df, gene, 'expression', 'mechanical_ventilation', 'sex') 
}
```





